Scalar Items in Embedded Position: {A}n Experimental Revisit

### 2012-09-01 text/paper/paper2.tex
Petra changes new version paper2

High scores of success in our general control condition clearly
indicate that participants understood the incremental verification
task and were in general able to give judgements at the earliest
possible point in a picture sequence. Results obtained for the
preference-related controls moreover show that our design is able to
detect multiple readings and preferences among these. Crucially, the
amount of response types for \lc- and \ec-readings reflected the known
preference for the former, \emph{irrespective} of the order in which
each reading could first be judged in a given sequence. Given this,
the best and most straightforward explanation of the observed data is
the hypothesis that the actual preferences among readings for both
\as- and \es-sentences alike are:
\begin{exe}
  \ex \label{bsp:observed-preferences} \lit \ > \ \loc \ \ ( > \glb)
\end{exe}
where it is unclear whether the global readings are entirely
unavailable or simply severely dispreferred. Crucially, our data
attests local readings independently of a stress accent on the
embedded scalar item, albeit that local readings are not preferred to
literal readings and given, as it seems, only by a small, but
substantial number of consistent localists.

This is bad news for both traditionalism and grammaticalism, as no
version of either would predict the pattern in
(\ref{bsp:observed-preferences}). The main problem is that
traditionalism, if left on its own, fails to predict local readings,
while grammaticalism, when it relies on the strongest meaning
hypothesis, fails to predict the right preference order. Let's look
more closely at each position in turn.

As for traditionalism, it is not a major obstacle for this position to
predict that global readings are available. The weak variety of
traditionalism holds that the extra assumptions needed to derive
global readings are not standardly available, and so traditionalism is
compatible with a lack of observed global readings. Alternatively,
traditionalists could argue that the task's rather high processing
demands lead to a suppression of global readings, in line with the
findings of \citet{NeysDe-NeysSchaeken2007:When-People-Are} that
working memory load negatively affects the number of scalar
implicature responses in sentence verification tasks. Unfortunately,
to the extent that global readings are explained away as dispreferred
or unavailable, traditionalism also looses the possibility of
accounting for local readings of \as-sentences in its unrestricted
variety.\fn{Remember that what we called unrestricted traditionalism
  can account for local readings of \as-sentences by assuming that a
  sentence of the form \emph{Some of the $X$ are related to some of
    the $Y$} is a feasible alternative.} This is because unrestricted
traditionalism explains local and global readings of \as-sentences in
essentially the same manner. On top of that, traditionalism cannot
explain local readings for \es-sentences. Since our design presented
sentence material auditorily, the traditionalist cannot resort to
effects of silent intonation to explain these local readings
away. Local readings showed on the neutral accentuation case as
well. Notice that it is not a problem for this argument that the local
responses were not emphasized with a contrastive stress on the
embedded scalar. In sum, traditionalism alone cannot explain
(\ref{bsp:observed-preferences}), but needs an additional explanation
of local readings for \emph{both} \as- \emph{and} \es-sentences.

Grammaticalism, on the other hand, fails to predict the preference
relation in (\ref{bsp:observed-preferences}), because these strictly
contradict the strongest meaning hypothesis in either of the two
varieties suggested by \citet{ChierchiaFox2008:The-Grammatical}. This
shows most clearly in the fact that both versions of the strongest
meaning hypothesis predict that local readings of \as-sentences are
preferred over literal ones. The problem, therefore, seems to be the
selection of readings by the strongest meaning hypothesis. In fact, if
(\ref{bsp:observed-preferences}) is the actual preference over
readings, then the strongest meaning hypothesis is also conceptually
on the wrong track, because no account that merely looks at the
logical strengths of readings can plausibly predict the preference
pattern in (\ref{bsp:observed-preferences}) for both \as- and
\es-sentences at the same time. (Try, if you please!)

It may seem as if grammaticalism is better suited to cope with the
data than traditionalism, because traditionalism cannot cope with
local readings for principled reasons while the only thing that
grammaticalism got wrong is the selection criterion over preferred
readings, i.e., it mistakenly put all its coin on the strongest
meaning hypothesis. Turning away from comparisons of logical strength,
a \emph{prima facie} plausible selection criterion that produces
(\ref{bsp:observed-preferences}) is ready at hand: if we assume that
each insertion of $\exh(\cdot)$ operator is \emph{costly}
anti-proportional to the depth of its embedding under other logical
operators, then we derive a preference of literal readings over local
ones over global ones. Whether such an \emph{economy principle} is
sound and in line with further empirical data would then be an
interesting matter of further investigation. On the entirely
speculative supposition that it did turn out to be a good predictor,
grammaticalism would be saved in a straightforward way. But is this an
argument for grammaticalism contra traditionalism? No it clearly is
not. The crucial point is that grammaticalism needs a selection
criterion, for without one it is not a predictive theory at all, and
that the only one that has been seriously proposed fails to predict
the observed data. That grammaticalism can do with many such selection
criteria is not, as such, an argument in its favor over
traditionalism. The case remains that both theories fail on the
observed data.

There is a variation of an existing theory of scalar implicature that
does explain the preference ordering (\ref{bsp:observed-preferences})
in a completely straightforward manner. The theory of interest is
\emph{conventionalism} (also sometimes called \emph{lexicalism})
\citep[e.g.][]{LevinsonPresumptiveMeanings2000,Chierchia:2004_ScalarImplicatures}.
The general idea is that if the upper-bounding inference from
\emph{some} to \emph{some but not all} occurs frequently, then it
would be inefficient to execute over and over again the Gricean
reasoning that traditionalists propose. Conventionalism therefore
assumes that scalar inferences have become routinized, hard-wired, so
much so that there are two lexical entries for a scalar item like
\emph{some}, one with logical meaning \emph{some and maybe all} and
one with the upper-bounded meaning \emph{some but not
  all}. \citet{LevinsonPresumptiveMeanings2000} proposed that the
latter is the preferred default meaning, a position which has
therefore ofttimes be addressed as \mymark{defaultism}. Unfortunately,
defaultism has strong empirical evidence against it, because several
processing studies show that the computation of scalar inferences
appears to be time- and effort-consuming, when they \emph{do} arise,
not when they \emph{don't}, as defaultism would have it
\citep[c.f.][]{BrehenyKatsos2006:Are-Generalised,BrehenyKatsos2008:Experimental-In,NeysDe-NeysSchaeken2007:When-People-Are}. But
there is also competing evidence that scalar inferences are available
in immediate, low-level processes, such as the resolution of
referential expressions in a visual-world paradigm
\citep{GrodnerKlein2010:Some-and-Possib}. For our purposes, it is
interesting to note that the preference relation in
(\ref{bsp:observed-preferences}) is predicted by the dual of
defaultism, call it \emph{literalism}, which posits a lexical
ambiguity but predicts that the logical meanings are \emph{ceteris
  paribus} preferred over the upper-bounded meanings.

@@

\noindent In the context of the present study, this would then be good
news for traditionalism, because a combined
traditionalism+conventionalism would predict the observed data
well. The combined position would even be able to explain something
that grammaticalism cannot make sense of as easily, namely the
observation that there were rather consistent localists among the
participants of our study. The combined tradionalism+conventionalism
could explain this by the assumption that some (lazy?, smart?)
speakers have made a conventionalized upper-bounding meaning entry for
scalar \emph{some} while others keep computing things on the fly. It
is not clear how grammaticalism can account for the intra-subjective
consistency. Still, this is not an argument \emph{for} traditionalism,
because the point remains that traditionalism as such fails to predict
the observed data.

@@

Several worries and objections are conceivable that might cast doubt
on the previous interpretation of our experimental data. In the
following we would like to address some of these that we consider most
pressing. Eventually, however, we conclude that the main point raised
above remains valid: neither traditionalism, nor grammaticalism
explains the observations even under alternative interpretations of the
data.

There are many ways in which it could be questioned whether our design
really reliably indicates the preference relation in
(\ref{bsp:observed-preferences}) among readings of our target
sentences. A first potential worry from that class that can be
dismissed rather quickly is that the incremental verification task
(\ivt) might simply select for the weakest possible reading. This idea
might be grounded in the observation that the preferred \lc-readings
of our target-related controls entail the dispreferred
\ec-readings. It could thus be argued that the difference in responses
for preference-related controls are due to the entailment properties,
not the known preference relation over \lc- and \ec-readings. But this
does not explain why the distribution of responses in the target
conditions are obviously not governed by logical strength. In fact, it
is a very noteworthy point that although entailment relations among
readings differ between \as- and \es-sentences, the observed response
pattern does not. (This is what is at the heart of the problem for a
grammaticalism that relies on the strongest meaning hypothesis.)

But maybe, a similar kind of objection goes, it is not feasible to
actually compare results from the preference-related controls to
results from our target conditions, because the kind of ``ambiguity''
is of a different kind. In the former we have a case of syntactic
ambiguity, in the latter a small number of competing pragmatic
enrichments. The former is a case of perceived ambiguity (some of our
subjects reported this in the
\emph{Nachbefragungsbogen}\dn{English?}), the latter arguably is
not \citep[c.f.][]{GeurtsPouscoulous2009:Embedded-Implic}. So perhaps
even if the distribution of responses is indicative of the known
preference patterns in preference-related controls, this would not
necessarily mean that the distribution of responses in target
conditions is indicative of preference relations.

>>>>

High scores of success clearly indicate that participants understood 
the incremental verification task and were in general able to give 
judgements at the earliest possible point in a picture sequence. 
Moreover, results obtained for the preference-related controls  
show that our design is able to detect multiple readings and preferences 
among these. Crucially, the amount of response types for \lc- and \ec-readings 
reflected the known preference for the former, \emph{irrespective} of the 
order in which each reading could first be judged in a given sequence. 
Finally, prosodic effects were restricted to the preference-related 
controls. As expected, the dispreferred \ec-reading was chosen 
significantly more often when supported by prosodic phrasing. Interestingly, 
contrastively accenting scalar \emph{some} had no effect on the observed 
decisions in \as- and \es-sentences.

Given the overall pattern of results, the most straightforward 
explanation of the observed data is the hypothesis that the actual 
preferences among readings for both \as- and \es-sentences alike are:

\begin{exe}
  \ex \label{bsp:observed-preferences} \lit \ > \ \loc \ \ ( > \glb)
\end{exe}

where it is unclear whether the global readings are entirely
unavailable or simply severely dispreferred. Crucially, our data
attests local readings independent of prosody, albeit that local 
readings are not preferred to literal readings and given, as it seems, 
only by a small, but substantial number of consistent localists.

These findings are hard to explain both by traditionalism and grammaticalism, 
as no version of either would predict the pattern in (\ref{bsp:observed-preferences}). 
Traditionalism, if left on its own, fails to predict local readings,
while grammaticalism, when it relies on the strongest meaning
hypothesis, fails to predict the right preference order. Let's look
more closely at each position in turn.

For traditionalism, the absence of global readings \textit{per se} can be
explained in a straightforward manner, for instance, by proposing
that the extra assumptions needed to derive global readings are not 
standardly available (as in the weak variety of traditionalism).
Alternatively, traditionalists could argue that the task's rather high 
processing demands lead to a suppression of global readings, in line with the
findings of \citet{NeysDe-NeysSchaeken2007:When-People-Are} that
working memory load negatively affects the number of scalar
implicature responses in sentence verification tasks. However, it then remains 
unclear why memory effects should be restricted to global readings.
As a result, approaches like the unrestricted variety of traditionalism thereby 
loose the possibility of accounting for local readings of \as-sentences.
\fn{Remember that what we called unrestricted traditionalism
  can account for local readings of \as-sentences by assuming that a
  sentence of the form \emph{Some of the $X$ are related to some of
    the $Y$} is a feasible alternative.} This is because unrestricted
traditionalism explains local and global readings of \as-sentences in
essentially the same manner. On top of that, traditionalism cannot
explain local readings for \es-sentences. Finally, since our materials were 
presented auditorily, we controlled for potential effects of silent
prosody. Importantly, local readings were observed in the accented as well
as in the unaccented condition, thus countering previous traditionalist
suggestions that local readings are to be attributed to specific pragmatic 
mechanisms.\dn{Petra: das habe ich hier so von der Intro übernommen. Sowohl
dort als auch hier fände ich es interessant, kurz zu sehen, was für alternative
pragmatische Mechanismen das sind - geht das in Kürze? Den nächsten Satz habe ich
in Bezug zum vorhergehenden Absatz nicht verstanden}

%showed on the neutral accentuation case as
%well. 
Notice that it is not a problem for this argument that the local
responses were not emphasized with a contrastive stress on the
embedded scalar. In sum, traditionalism alone cannot explain
(\ref{bsp:observed-preferences}), but needs an additional explanation
of local readings for \emph{both} \as- \emph{and} \es-sentences.

Grammaticalism, on the other hand, fails to predict the preference
relation in (\ref{bsp:observed-preferences}), because it strictly
contradicts the strongest meaning hypothesis in either of the two
varieties suggested by \citet{ChierchiaFox2008:The-Grammatical}. This
shows most clearly in the fact that both versions of the strongest
meaning hypothesis predict that local readings of \as-sentences are
preferred over literal ones. The problem, therefore, seems to be the
selection of readings by the strongest meaning hypothesis. In fact, if
(\ref{bsp:observed-preferences}) is the actual preference over
readings, then the strongest meaning hypothesis is also conceptually
on the wrong track, because no account that merely looks at the
logical strengths of readings can plausibly predict the preference
pattern in (\ref{bsp:observed-preferences}) for both \as- and
\es-sentences at the same time. (Try, if you please!)

Summing up so far, the existing versions of traditionalism and grammaticalism
cannot explain the present results. Of the two approaches, grammaticalism 
seems better suited to account for the observed effects. Whereas 
traditionalism cannot cope with local readings for principled reasons, 
the only thing that grammaticalism would have to change is the 
selection criterion over preferred readings, i.e. to reconsider the strongest
meaning hypothesis. Turning away from comparisons of logical strength,
a \emph{prima facie} plausible selection criterion that produces
(\ref{bsp:observed-preferences}) is ready at hand: if we assume that
each insertion of $\exh(\cdot)$ operator is \emph{costly}
anti-proportional to the depth of its embedding under other logical
operators, then we derive a preference of literal readings over local
ones over global ones. Whether such an \emph{economy principle} is
sound and in line with further empirical data would then be an
interesting matter of further investigation. 
%On the entirely
%speculative supposition that it did turn out to be a good predictor,
%grammaticalism would be saved in a straightforward way. 
%But is this an argument for grammaticalism contra traditionalism? No 
%it clearly is not.

%In order to be a predictive theory, grammaticalism thus needs a selection
%criterion, for without one it is not a predictive theory at all, and
%that the only one that has been seriously proposed fails to predict
%the observed data. That grammaticalism can do with many such selection
%criteria is not, as such, an argument in its favor over
%traditionalism. The case remains that both theories fail on the
%observed data.

Interestingly, an alternative theory of scalar implicature explains the 
observed preference ordering (\ref{bsp:observed-preferences})
in a completely straightforward manner. 
\emph{Conventionalism} (also sometimes called \emph{lexicalism})
\citep[e.g.][]{LevinsonPresumptiveMeanings2000,Chierchia:2004_ScalarImplicatures}.
proposes that if the upper-bounding inference from
\emph{some} to \emph{some but not all} occurs frequently, it
would be inefficient to execute over and over again the Gricean
reasoning that traditionalists propose. Conventionalism therefore
assumes that scalar inferences have become routinized, hard-wired, so
much so that there are two lexical entries for a scalar item like
\emph{some}, one with logical meaning \emph{some and maybe all} and
one with the upper-bounded meaning \emph{some but not
  all}. \citet{LevinsonPresumptiveMeanings2000} proposed that the
latter is the preferred default meaning, a position which has
therefore ofttimes be addressed as \mymark{defaultism}. Unfortunately,
defaultism has strong empirical evidence against it, because several
processing studies show that the computation of scalar inferences
appears to be time- and effort-consuming, when they \emph{do} arise,
not when they \emph{don't}, as defaultism would have it
\citep[c.f.][]{BrehenyKatsos2006:Are-Generalised,BrehenyKatsos2008:Experimental-In,NeysDe-NeysSchaeken2007:When-People-Are}. But
there is also competing evidence that scalar inferences are available
in immediate, low-level processes, such as the resolution of
referential expressions in a visual-world paradigm
\citep{GrodnerKlein2010:Some-and-Possib}. For our purposes, it is
interesting to note that the preference relation in
(\ref{bsp:observed-preferences}) is predicted by the dual of
defaultism, call it \emph{literalism}, which posits a lexical
ambiguity but predicts that the logical meanings are \emph{ceteris
  paribus} preferred over the upper-bounded meanings.

@@

\noindent In the context of the present study, this would then be good
news for traditionalism, because a combined
traditionalism+conventionalism would predict the observed data
well. The combined position would even be able to explain something
that grammaticalism cannot make sense of as easily, namely the
observation that there were rather consistent localists among the
participants of our study. The combined tradionalism+conventionalism
could explain this by the assumption that some (lazy?, smart?)
speakers have made a conventionalized upper-bounding meaning entry for
scalar \emph{some} while others keep computing things on the fly. It
is not clear how grammaticalism can account for the intra-subjective
consistency. Still, this is not an argument \emph{for} traditionalism,
because the point remains that traditionalism as such fails to predict
the observed data.\dn{Petra: dieser Absatz scheint mir sehr wichtig und gut - 
wäre es möglich, hier noch etwas ausführlicher drüber zu schreiben?}

@@

%Several worries and objections are conceivable that might cast doubt
%on the previous interpretation of our experimental data.

In the following, we would like to address potential alternative
explanations for the present results. We will argue that besides
these interpretation variants, the main point raised
above remains valid: neither traditionalism, nor grammaticalism
is able to account for the present pattern of results.

%some of these that we consider most
%pressing. Eventually, however, we conclude that the main point raised
%above remains valid: neither traditionalism, nor grammaticalism
%explains the observations even under alternative interpretations of the
%data.

First, it could be questioned whether our design reliably 
indicates the preference relation in (\ref{bsp:observed-preferences}) 
among readings of our target sentences. Thus, the incremental verification task
(\ivt) might simply select for the weakest possible reading. This idea
can be grounded in the observation that the preferred \lc-readings
of our target-related controls entail the dispreferred
\ec-readings. It could thus be argued that the difference in responses
for preference-related controls are due to the entailment properties,
not the known preference relation over \lc- and \ec-readings. However, this
does not explain why the distribution of responses in the target
conditions are obviously not governed by logical strength. In fact, it
is a very noteworthy point that although entailment relations among
readings differ between \as- and \es-sentences, the observed response
pattern does not. (This is what is at the heart of the problem for a
grammaticalism that relies on the strongest meaning hypothesis.)
\dn{Petra: ich glaub, den Punkt verstehe ich nicht ganz - sollte man hier
auch die Prosodie erwähnen? Phrasierung wirkt sich ja nicht auf entailment
relations aus, sondern wird normalerweise dem Syntax-Prosodie-Interface
zugeschrieben}

But maybe, a similar kind of objection goes, it is not feasible to
actually compare results from the preference-related controls to
results from our target conditions, because the kind of ``ambiguity''
is of a different kind. In the former we have a syntactic
ambiguity, in the latter a small number of competing pragmatic
enrichments. The former is a case of perceived ambiguity (some of our
subjects reported this in a \emph{Nachbefragungsbogen}\dn{English?}), 
the latter arguably is
not \citep[c.f.][]{GeurtsPouscoulous2009:Embedded-Implic}. So perhaps
even if the distribution of responses is indicative of the known
preference patterns in preference-related controls, this would not
necessarily mean that the distribution of responses in target
conditions is indicative of preference relations.
